---
title: "Demo 28: What the Data Showed"
subtitle: "DSC 011 Sp26 · Submission Analysis (anonymized) · n = 52 (Linux/UNIX Project Building and Infrastructure Check)"
author: "David H. Ardell with Claude Sonnet 4.6"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    theme: cosmo
    embed-resources: true
    fig-width: 9
    fig-height: 5.5
execute:
  echo: true
  warning: false
  message: false
engine: knitr
---

```{r setup, include=FALSE}
options(digits = 5)
knitr::opts_chunk$set(
  echo    = TRUE,
  cache   = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = "",
  fig.align = "center"
)
```

```{r packages, include=FALSE}
# install.packages(c("ggplot2","dplyr","tidyr","forcats","scales","stringr"))
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(scales)
library(stringr)
```

```{r palette, include=FALSE}
OI <- c(
  orange = "#E69F00", sky_blue = "#56B4E9", green = "#009E73",
  yellow = "#F0E442", blue = "#0072B2", vermillion = "#D55E00",
  pink = "#CC79A7",   black = "#000000"
)
clr_correct  <- OI[["green"]];  clr_incorrect <- OI[["orange"]]
clr_err      <- OI[["vermillion"]]; clr_noout  <- OI[["sky_blue"]]
clr_macos_good <- OI[["blue"]]; clr_wsl  <- OI[["green"]]
clr_macos_old  <- OI[["pink"]]; clr_crlf <- OI[["orange"]]

result_colours <- c(
  "CORRECT" = clr_correct, "INCORRECT" = clr_incorrect,
  "Error running bash" = clr_err, "No output / not run" = clr_noout
)
plat_colours <- c(
  "macOS (Homebrew bash 5)" = clr_macos_good,
  "WSL2/Linux (bash 5.2)"   = clr_wsl,
  "macOS (/bin/bash 3.2, unfixed)" = clr_macos_old,
  "WSL2 (CRLF line endings)" = clr_crlf
)
cwd_colours <- c(
  "Correct" = clr_correct,
  "Windows path (/mnt/c/ or /c/)" = clr_err,
  "Downloads/" = clr_incorrect,
  "Desktop/"   = clr_crlf,
  "Other wrong location" = clr_noout
)
theme_dsc <- function(base = 13) {
  theme_minimal(base_size = base) +
    theme(
      plot.title   = element_text(face = "bold", size = base + 1, margin = margin(b = 4)),
      plot.subtitle = element_text(size = base - 1, color = "grey30", lineheight = 1.3,
                                   margin = margin(b = 6)),
      plot.caption = element_text(size = base - 3, color = "grey50", hjust = 0,
                                  lineheight = 1.35, margin = margin(t = 8)),
      legend.position = "bottom", legend.title = element_blank(),
      legend.text = element_text(size = base - 2),
      panel.grid.minor = element_blank(), plot.margin = margin(6, 6, 6, 6)
    )
}
```

```{r data, include=FALSE}
raw <- tibble(
  student = paste0("student_", 1:52),
  late = c(
    TRUE, FALSE, FALSE, TRUE, FALSE, TRUE, TRUE, FALSE, TRUE, FALSE,
    TRUE, FALSE, TRUE, FALSE, TRUE, FALSE, FALSE, TRUE, FALSE, TRUE,
    FALSE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE, TRUE, FALSE, FALSE,
    TRUE, TRUE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, FALSE, TRUE,
    FALSE, TRUE, FALSE, FALSE, FALSE, TRUE, FALSE, TRUE, FALSE, TRUE,
    FALSE, TRUE
  ),
  platform = c(
    "macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)",
    "WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)",
    "macOS (Homebrew bash 5)","macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)",
    "macOS (Homebrew bash 5)","WSL2 (CRLF line endings)","macOS (Homebrew bash 5)",
    "macOS (/bin/bash 3.2, unfixed)","WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)",
    "WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)",
    "macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)","macOS (/bin/bash 3.2, unfixed)",
    "WSL2 (CRLF line endings)","macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)",
    "macOS (/bin/bash 3.2, unfixed)","WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)",
    "WSL2/Linux (bash 5.2)","WSL2 (CRLF line endings)","macOS (Homebrew bash 5)",
    "macOS (Homebrew bash 5)","macOS (/bin/bash 3.2, unfixed)","WSL2/Linux (bash 5.2)",
    "macOS (Homebrew bash 5)","macOS (Homebrew bash 5)","WSL2 (CRLF line endings)",
    "WSL2/Linux (bash 5.2)","macOS (/bin/bash 3.2, unfixed)","macOS (Homebrew bash 5)",
    "WSL2/Linux (bash 5.2)","WSL2 (CRLF line endings)","macOS (Homebrew bash 5)",
    "WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)",
    "macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)",
    "WSL2/Linux (bash 5.2)","macOS (/bin/bash 3.2, unfixed)","WSL2/Linux (bash 5.2)",
    "WSL2/Linux (bash 5.2)"
  ),
  correct_cwd = c(
    TRUE,FALSE,TRUE,TRUE,TRUE,FALSE,TRUE,FALSE,TRUE,FALSE,FALSE,FALSE,FALSE,TRUE,
    TRUE,FALSE,FALSE,TRUE,FALSE,TRUE,FALSE,FALSE,FALSE,TRUE,FALSE,FALSE,TRUE,TRUE,
    FALSE,FALSE,TRUE,FALSE,TRUE,FALSE,FALSE,TRUE,TRUE,FALSE,TRUE,FALSE,FALSE,TRUE,
    TRUE,TRUE,FALSE,TRUE,FALSE,FALSE,TRUE,FALSE,TRUE,FALSE
  ),
  cwd_category = c(
    "Correct","Desktop/","Correct","Correct","Correct","Other wrong location","Correct",
    "Windows path (/mnt/c/ or /c/)","Correct","Downloads/",
    "Windows path (/mnt/c/ or /c/)","Desktop/","Downloads/","Correct","Desktop/",
    "Correct","Downloads/","Correct","Downloads/","Correct","Other wrong location",
    "Windows path (/mnt/c/ or /c/)","Other wrong location","Correct","Downloads/",
    "Correct","Correct","Correct","Windows path (/mnt/c/ or /c/)","Downloads/",
    "Other wrong location","Downloads/","Correct","Downloads/","Desktop/",
    "Windows path (/mnt/c/ or /c/)","Correct","Other wrong location","Correct",
    "Other wrong location","Windows path (/mnt/c/ or /c/)","Correct",
    "Other wrong location","Desktop/","Other wrong location","Correct",
    "Other wrong location","Downloads/","Correct","Windows path (/mnt/c/ or /c/)",
    "Other wrong location","Windows path (/mnt/c/ or /c/)"
  ),
  export_fix = c(
    FALSE,FALSE,FALSE,TRUE,FALSE,FALSE,FALSE,FALSE,TRUE,FALSE,FALSE,FALSE,FALSE,
    FALSE,FALSE,TRUE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,
    FALSE,FALSE,FALSE,FALSE,TRUE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,
    TRUE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,TRUE,FALSE
  ),
  cwd_display = c(
    "~/DSC011/DSC-011-Sp26-Baby-Names","/Users/student_2/Desktop/DSC_011_S26",
    "~/DSC011/DSC-011-Sp26-Baby-Names","~/DSC011/DSC-011-Sp26-Baby-Names",
    "~/DSC011/DSC-011-Sp26-Baby-Names","/home/student_6/DSC011/DEMOS",
    "~/DSC011/DSC-011-Sp26-Baby-Names","/c/Users/student_8/OneDrive/DEMOS",
    "~/DSC011/DSC-011-Sp26-Baby-Names","/Users/student_10/Downloads",
    "/mnt/c/Users/student_11/Downloads","/Users/student_12/Desktop",
    "/Users/student_13/Downloads","~/DSC011/DSC-011-Sp26-Baby-Names",
    "~/Desktop/DSC_011_S26/DSC-011-Sp26-Baby-Names","~/DSC011/DSC-011-Sp26-Baby-Names",
    "/Users/student_17/Downloads","~/DSC011/DSC-011-Sp26-Baby-Names",
    "/Users/student_19/Downloads","~/DSC011/DSC-011-Sp26-Baby-Names",
    "/Users/student_21/Documents/DSC 011/Lecture28_Demo",
    "/mnt/c/Users/student_22/OneDrive/Documents/DSC_011_S26/DEMOS",
    "/Users/student_23/DCS_011_026/demos","~/DSC011/DSC-011-Sp26-Baby-Names",
    "/Users/student_25/Downloads","~/DSC011/DSC-011-Sp26-Baby-Names",
    "~/DSC011/DSC-011-Sp26-Baby-Names","~/DSC011/DSC-011-Sp26-Baby-Names",
    "/mnt/c/Users/student_29/Downloads","/Users/student_30/Downloads",
    "~/MyStuff/DSC_011_S26/DSC-011-Sp26-Baby-Names","/Users/student_32/Downloads",
    "~/DSC011/DSC-011-Sp26-Baby-Names","/Users/student_34/Downloads",
    "/Users/student_35/Desktop/DSC_011_S26/DEMOS",
    "/mnt/c/Users/student_36/OneDrive/Documents/DSC_011_S26/DEMOS",
    "~/DSC011/DSC-011-Sp26-Baby-Names","/Users/student_38/Downloads/DSC011-S26 class/DEMOS",
    "~/DSC011/DSC-011-Sp26-Baby-Names","~/DSC011/Lecture28/DSC-011-Sp26-Baby-Names",
    "/mnt/c/Users/student_41/Downloads","~/DSC011/DSC-011-Sp26-Baby-Names",
    "~/DSC-011/DSC-011-Sp26-Baby-Names","~/Desktop/Demos/DSC-011-Sp26-Baby-Names",
    "~/DSC011","/home/student_47/DSC011 wait no~/DSC011/DSC-011-Sp26-Baby-Names",
    "~/DSC011/DSC-011-Sp26-Baby-Names","/Users/student_48/Downloads",
    "~/DSC011/DSC-011-Sp26-Baby-Names",
    "/mnt/c/Users/student_50/OneDrive/Spring 2026/DSC_011_S26/DEMOS",
    "~/dsc/DSC-011-Sp26-Baby-Names","/mnt/c/Users/student_52/Downloads"
  ),
  dir_scaffold = c(
    "ERR","ERR","CORRECT","ERR","CORRECT","ERR","no_output","no_output","no_output",
    "no_output","no_output","no_output","ERR","no_output","INCORRECT","ERR","ERR",
    "no_output","ERR","no_output","ERR","no_output","INCORRECT","ERR","ERR","ERR",
    "no_output","no_output","no_output","no_output","no_output","ERR","ERR","ERR",
    "ERR","no_output","ERR","ERR","no_output","ERR","no_output","CORRECT","no_output",
    "ERR","ERR","no_output","ERR","ERR","ERR","no_output","no_output","no_output"
  ),
  python_env = c(
    "ERR","ERR","CORRECT","ERR","ERR","ERR","INCORRECT","no_output","ERR","no_output",
    "no_output","no_output","ERR","ERR","INCORRECT","ERR","ERR","ERR","ERR","no_output",
    "ERR","no_output","INCORRECT","ERR","ERR","ERR","CORRECT","ERR","no_output",
    "no_output","ERR","ERR","ERR","ERR","ERR","no_output","ERR","ERR","CORRECT","ERR",
    "no_output","CORRECT","ERR","ERR","ERR","no_output","ERR","ERR","ERR","no_output",
    "ERR","no_output"
  ),
  setup_sh = c(
    "ERR","ERR","CORRECT","ERR","ERR","ERR","INCORRECT","no_output","ERR","no_output",
    "no_output","INCORRECT","ERR","ERR","INCORRECT","ERR","ERR","ERR","ERR","INCORRECT",
    "ERR","no_output","INCORRECT","ERR","ERR","ERR","ERR","ERR","no_output","no_output",
    "ERR","ERR","ERR","ERR","ERR","no_output","ERR","ERR","CORRECT","ERR","no_output",
    "INCORRECT","ERR","ERR","ERR","INCORRECT","ERR","ERR","ERR","no_output","ERR",
    "no_output"
  ),
  git_commit_1 = c(
    "ERR","ERR","INCORRECT","ERR","ERR","ERR","INCORRECT","no_output","ERR","no_output",
    "no_output","INCORRECT","ERR","ERR","INCORRECT","ERR","ERR","ERR","ERR","INCORRECT",
    "ERR","no_output","INCORRECT","ERR","ERR","ERR","ERR","ERR","no_output","no_output",
    "ERR","ERR","ERR","ERR","ERR","no_output","ERR","ERR","INCORRECT","ERR","no_output",
    "INCORRECT","ERR","ERR","ERR","INCORRECT","ERR","ERR","ERR","no_output","ERR",
    "no_output"
  ),
  html_14_files = c(
    "ERR","ERR","INCORRECT","ERR","ERR","ERR","INCORRECT","no_output","ERR","no_output",
    "no_output","INCORRECT","ERR","ERR","INCORRECT","ERR","ERR","ERR","ERR","INCORRECT",
    "ERR","no_output","INCORRECT","ERR","ERR","ERR","ERR","ERR","no_output","no_output",
    "ERR","ERR","ERR","ERR","ERR","no_output","ERR","ERR","CORRECT","ERR","no_output",
    "INCORRECT","ERR","ERR","ERR","INCORRECT","ERR","ERR","ERR","no_output","ERR",
    "no_output"
  ),
  git_commit_2 = c(
    "ERR","ERR","INCORRECT","ERR","ERR","ERR","INCORRECT","no_output","ERR","no_output",
    "no_output","INCORRECT","ERR","ERR","INCORRECT","ERR","ERR","ERR","ERR","INCORRECT",
    "ERR","no_output","INCORRECT","ERR","ERR","ERR","ERR","ERR","no_output","no_output",
    "ERR","ERR","ERR","ERR","ERR","no_output","ERR","ERR","CORRECT","ERR","no_output",
    "CORRECT","ERR","ERR","ERR","INCORRECT","ERR","ERR","ERR","no_output","ERR",
    "no_output"
  )
)

n_total <- nrow(raw)
chunk_levels <- c("dir_scaffold","python_env","setup_sh",
                  "git_commit_1","html_14_files","git_commit_2")
chunk_labels <- c(
  dir_scaffold = "Dir scaffold", python_env = "Python env",
  setup_sh = "00-setup.sh", git_commit_1 = "Git commits = 1",
  html_14_files = "14 HTML files", git_commit_2 = "Git commits = 2"
)
result_levels <- c("CORRECT","INCORRECT","ERR","no_output")
result_labels  <- c("CORRECT","INCORRECT","Error running bash","No output / not run")
plat_levels <- c("macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)",
                 "macOS (/bin/bash 3.2, unfixed)","WSL2 (CRLF line endings)")
cwd_cat_levels <- c("Correct","Windows path (/mnt/c/ or /c/)",
                    "Downloads/","Desktop/","Other wrong location")

students <- raw |>
  mutate(
    platform     = factor(platform,     levels = plat_levels),
    cwd_category = factor(cwd_category, levels = cwd_cat_levels),
    submission   = if_else(late, "Late", "On time")
  )

n_windows_path <- sum(students$cwd_category == "Windows path (/mnt/c/ or /c/)")
n_downloads    <- sum(students$cwd_category == "Downloads/")
n_desktop      <- sum(students$cwd_category == "Desktop/")
n_other_wrong  <- sum(students$cwd_category == "Other wrong location")
n_correct_cwd  <- sum(students$correct_cwd)
n_wrong_cwd    <- n_total - n_correct_cwd
```

## About This Document

This is a **reproducible data analysis deliverable** — the complete source
code, anonymized data, and interpretive prose needed for any reader to
understand, verify, and reproduce every figure and conclusion. It is shared
both as a transparency measure and as a model of the kind of analysis product
this course is building toward.

**What was analyzed.** All `r n_total` rendered HTML submissions from
Demo 28 (DSC 011 Sp26, April 7 2026) were parsed automatically. Each HTML
file was stripped of markup tags using `sed` and concatenated into a single
text file. A Python script then split the file on student-boundary markers
and extracted: the bash binary path knitr used (a platform proxy), the
working directory at render time, and the outcome of each of the six
`check_bash` validation chunks in the notebook.

**Why it is being shared.** Three independent failure modes accounted for
nearly all Demo 28 errors. Each has been fixed in Demo 29 and Lab 10.
Understanding *why* the fixes were made — not just *what* changed — is part
of the course's learning goals.

**Anonymization.** All student names are replaced with randomized IDs
(`student_1` through `student_52`). The shuffle order is arbitrary. Working
directory paths retain the structure of the error (e.g., `Downloads/`,
`OneDrive/`) but real usernames are replaced with `student_N` tokens.

## Anonymization Validation

The chunk below sweeps every character column for strings outside a
known-safe category list. All columns should read **OK** before distribution.

```{r id-validation}
known_safe <- c(
  "macOS","WSL2","Linux","bash","CRLF","unfixed","student_",
  "CORRECT","INCORRECT","ERR","no_output","DSC","Downloads","Desktop",
  "OneDrive","Documents","DEMOS","Lecture","home","Users","mnt","dsc",
  "MyStuff","Spring","class","Correct","Windows","path","Other","wrong",
  "location","Homebrew","wait","no"
)
id_check <- students |>
  select(student, platform, cwd_display, cwd_category,
         dir_scaffold, python_env, setup_sh,
         git_commit_1, html_14_files, git_commit_2) |>
  summarise(across(where(is.character), \(col) {
    flagged <- col[!str_detect(col, paste(known_safe, collapse = "|"))]
    if (length(flagged) == 0) "OK" else paste("FLAGGED:", paste(flagged, collapse = "; "))
  })) |>
  pivot_longer(everything(), names_to = "Column", values_to = "Result")
knitr::kable(id_check, align = c("l","l"))
```

## How the Data Were Collected

Submissions were collected as rendered HTML files. Because HTML files cannot
be uploaded individually in large batches, all files were stripped and
concatenated in the terminal:

```bash
for f in *.html; do
    echo "=== STUDENT: $f ==="
    sed 's/<[^>]*>//g' "$f"
done > all_submissions_stripped.txt
```

For each student the parser extracted:

- **Platform** — inferred from the bash binary path in the bash version
  guard chunk output (`✓ /opt/homebrew/bin/bash` → macOS Homebrew; `✓
  /usr/bin/bash` → WSL2). Students whose guard produced no output required
  secondary inference.
- **Working directory** — from `pwd` output where present.
- **Chunk outcomes** — matched by the SHA-256 hash in each `check_bash` call,
  then read from the following lines for `CORRECT`, `INCORRECT`, or the
  `Error in running command bash` knitr error string.
- **Workaround use** — detected by `export PATH=` immediately before
  `answer=` in the same chunk body.

## The Four Platform Groups

Every Demo 28 notebook printed the bash binary path and version when the
version guard chunk ran. This path reliably identifies the student's platform
configuration. The class uses two operating systems, but each produced two
distinct sub-groups — giving four groups total.

**macOS (Homebrew bash 5):** Homebrew installed, RStudio launched from the
terminal. knitr used `/opt/homebrew/bin/bash` version 5.3.9.

**WSL2/Linux (bash 5.2):** Correctly configured WSL2 environment. knitr
used `/usr/bin/bash` version 5.2.21.

**macOS (/bin/bash 3.2, unfixed):** RStudio launched from the Dock.
macOS's process manager (`launchd`) gives GUI apps a minimal PATH that
excludes `/opt/homebrew/bin/`, so knitr fell back to Apple's GPL-frozen
`/bin/bash` 3.2 from 2007. The bash version guard was designed to catch
this, but Demo 28 used `error=TRUE` (not `error=FALSE`), so knitr swallowed
the non-zero exit silently. All subsequent chunks ran with a broken
environment.

**WSL2 (CRLF line endings):** These students downloaded the `.qmd` file
through a Windows browser, which saved it with `\r\n` line endings. Inside
WSL2's Linux kernel, bash reads the `\r` as part of each command name,
producing `cannot: command not found` for every chunk. No bash command could
execute. This is a completely distinct failure mode from the PATH problem.

```{r fig-platform, echo=FALSE}
fig1_data <- students |> count(platform, submission)
ggplot(fig1_data, aes(x = submission, y = n, fill = platform)) +
  geom_col(width = 0.45, colour = "white", linewidth = 0.6) +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            size = 5, fontface = "bold", colour = "white") +
  scale_fill_manual(values = plat_colours) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
  labs(title = "Two platform groups worked as expected; two reflect misconfiguration",
       subtitle = paste0(
         "macOS /bin/bash 3.2: RStudio launched from Dock  ·  ",
         "WSL2 CRLF: .qmd downloaded through Windows browser"),
       x = NULL, y = "Students",
       caption = paste0("n = ", n_total,
                        "  ·  22 macOS Homebrew, 18 WSL2 Linux, 6 macOS unfixed, 6 WSL2 CRLF")) +
  theme_dsc()
```

## Failure Mode 1 — Wrong Working Directory

When Quarto renders a notebook, knitr sets the working directory to the
folder containing the `.qmd` file. Every bash chunk runs with that folder
as its current directory. Demo 28 was designed to live at the project root
(`~/DSC011/DSC-011-Sp26-Baby-Names/`) so that relative paths like `data/raw/`
and `scripts/` would resolve. If the notebook was anywhere else, those paths
could not be found.

`r n_wrong_cwd` of `r n_total` students (`r round(n_wrong_cwd/n_total*100)`%)
rendered from the wrong directory. The breakdown:

- **`r n_windows_path` students (15%) — Windows filesystem path** (`/mnt/c/`
  or `/c/`). These students were working on the Windows side of the WSL2
  boundary. This is discussed further below.
- **`r n_downloads` students (19%) — `Downloads/`**. The notebook was never
  moved from the browser's download location.
- **`r n_desktop` students (10%) — `Desktop/`**. The notebook was placed on
  the desktop, not in the course-standard location.
- **`r n_other_wrong` students (17%) — Other wrong location**. Included:
  wrong project name (`DSC-011` vs `DSC011`, `DCS_011_026` vs `DSC_011`),
  notebook nested inside a `Lecture28/` subfolder, project at the parent
  directory rather than the project root, and custom non-standard paths.

**Why exact names matter.** Directory and file names in this course are part
of the interface between the notebook, the shell scripts, and the validation
checks. A path differing by one character is a different path. Directory and
file names should never be typed by hand — use copy-paste from course
materials or tab-completion in the terminal.

```{r fig-cwd-breakdown, echo=FALSE}
cwd_data <- students |> count(cwd_category) |> mutate(cwd_category = fct_rev(cwd_category))
ggplot(cwd_data, aes(x = cwd_category, y = n, fill = cwd_category)) +
  geom_col(width = 0.55, colour = "white", linewidth = 0.5) +
  geom_text(aes(label = paste0(n, " (", round(n/n_total*100), "%)")),
            hjust = -0.1, size = 4.2, fontface = "bold", colour = "grey20") +
  scale_fill_manual(values = cwd_colours, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.3))) +
  coord_flip() +
  labs(title = paste0(n_wrong_cwd, " of ", n_total, " students (",
                      round(n_wrong_cwd/n_total*100), "%) rendered from the wrong directory"),
       x = NULL, y = "Students",
       caption = "Demo 29 fix: error=FALSE CWD guard halts rendering if path ≠ expected.") +
  theme_dsc()
```

### Working Directory Examples

```{r cwd-table, echo=FALSE}
wrong_cwds <- students |>
  filter(!correct_cwd) |>
  select(Student = student, Platform = platform,
         Category = cwd_category, `Working directory` = cwd_display) |>
  arrange(Category, Student)
knitr::kable(wrong_cwds, align = c("l","l","l","l"))
```

### Windows Filesystem Paths Are Not Viable

`r n_windows_path` of `r n_total` students (15%) were working from `/mnt/c/`
or `/c/` paths — the Windows NTFS filesystem mounted inside WSL2.

WSL2 provides two separate filesystems: the Linux filesystem (`~/`, stored
inside the virtual machine) and the Windows filesystem (`/mnt/c/`, on the
physical Windows drive). Files can be read from either, but they are not
equivalent environments.

Working from `/mnt/c/` causes: CRLF line ending contamination (files
written by Windows tools get `\r\n`); significantly slower filesystem
operations due to cross-boundary overhead; unreliable `chmod` and executable
bits; symlink behavior differences; and paths that are specific to one
machine's Windows configuration and cannot be reproduced or shared.

**All course work lives in `~/DSC-011-S26/` on the Linux filesystem.** If
the terminal prompt shows a path beginning with `/mnt/c/`, the working
location is wrong.

## Failure Mode 2 — `check_bash` Not on PATH

In Demo 28, `check_bash` was installed to `~/bin/` via a curl one-liner,
which also added `export PATH="$HOME/bin:$PATH"` to the student's shell
startup file (`.bashrc` or `.zshrc`). This works correctly in an interactive
terminal session, where the startup file is sourced each time a new shell
opens.

**How knitr runs bash chunks.** Each bash chunk is a fresh, *non-interactive*
bash subprocess. knitr writes the chunk body to a temporary file and calls
`bash <tempfile>`. Non-interactive bash does not source `.bashrc` or `.zshrc`.
It inherits only the environment R has at render time, which comes from
whatever process launched RStudio. Neither macOS GUI-launched RStudio nor
WSL2 RStudio Server sources user startup files before spawning knitr
subprocesses.

The result: `check_bash` was installed correctly and worked in the terminal.
But every knitr bash chunk calling `check_bash` received
`bash: check_bash: command not found` (exit code 127). knitr renders a
non-zero bash exit as `Error in running command bash` — the error seen in
63% of submissions.

**Why the guard chunk still passed.** The bash version guard used only bash
builtins (`BASH_VERSINFO`, `echo`, `which`) — no `check_bash`. So it passed,
giving false confidence before every subsequent chunk failed.

**The correct fix.** Install software that must be available to all processes
— including non-interactive subprocesses — into a system-level directory.
`/usr/local/bin/` is on the default PATH of every bash subprocess on both
macOS and Linux, regardless of how RStudio was launched. This principle
applies beyond `check_bash`: any tool needed inside knitr chunks should be
installed system-wide (via `sudo`) or via an environment manager like `pixi`
that places binaries on the activated PATH.

```{r fig-chunks, echo=FALSE}
fig3_data <- students |>
  select(student, all_of(chunk_levels)) |>
  pivot_longer(-student, names_to = "chunk", values_to = "result") |>
  mutate(chunk  = factor(chunk,  levels = chunk_levels, labels = chunk_labels),
         result = factor(result, levels = result_levels, labels = result_labels)) |>
  count(chunk, result)
ggplot(fig3_data, aes(x = chunk, y = n, fill = result)) +
  geom_col(width = 0.55, colour = "white", linewidth = 0.5) +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            size = 4.8, fontface = "bold", colour = "white") +
  scale_fill_manual(values = result_colours) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.04)), breaks = seq(0, 52, 10)) +
  labs(title = "0 students passed dir scaffold · 33 had bash errors on every chunk",
       subtitle = paste0(
         "'Error running bash' = check_bash not found (exit 127)  ·  ",
         "'No output' = chunk skipped or CRLF broken  ·  ",
         "'INCORRECT' = check_bash ran, answer was wrong"),
       x = NULL, y = "Students",
       caption = "Demo 29 fix: check_bash 2.2 installed to /usr/local/bin/ — on PATH everywhere.") +
  theme_dsc() + theme(axis.text.x = element_text(size = 10.5))
```

### The PATH Error Was Platform-Independent

The error affected macOS and WSL2 at nearly identical rates, confirming this
was not a macOS GUI PATH issue — it was architectural.

```{r fig-heatmap, echo=FALSE, fig.height=4}
fig4_data <- students |>
  filter(platform %in% c("macOS (Homebrew bash 5)", "WSL2/Linux (bash 5.2)")) |>
  select(student, platform, all_of(chunk_levels)) |>
  pivot_longer(all_of(chunk_levels), names_to = "chunk", values_to = "result") |>
  mutate(chunk = factor(chunk, levels = chunk_levels, labels = chunk_labels),
         is_err = result == "ERR") |>
  group_by(platform, chunk) |>
  summarise(n_err = sum(is_err), n = n(), pct = n_err / n, .groups = "drop")
ggplot(fig4_data, aes(x = chunk, y = fct_rev(platform), fill = pct)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = paste0(n_err, "/", n, " (", round(pct*100), "%)")), size = 3.8) +
  scale_fill_gradient(low = "#fffde7", high = clr_err,
                      labels = percent_format(accuracy = 1),
                      name = "% with error", limits = c(0,1)) +
  labs(title = "macOS and WSL2 students failed at nearly identical rates",
       subtitle = "Non-interactive bash does not source .bashrc on either platform.",
       x = NULL, y = NULL,
       caption = "Fix: /usr/local/bin/ is on the default PATH of every bash subprocess on both platforms.") +
  theme_dsc() + theme(panel.grid = element_blank(), legend.position = "right")
```

### The Per-Chunk Workaround

Six students added `export PATH="$HOME/bin:$PATH"` to individual chunks —
a correct diagnosis but inconsistently applied, so most chunks still errored.

```{r fig-workaround, echo=FALSE, fig.height=4.5}
fig5_data <- students |>
  mutate(workaround = if_else(export_fix,
    paste0("Used export PATH workaround (n=", sum(export_fix), ")"),
    paste0("Did not (n=", sum(!export_fix), ")"))) |>
  select(student, workaround, all_of(chunk_levels)) |>
  pivot_longer(all_of(chunk_levels), names_to = "chunk", values_to = "result") |>
  mutate(result = factor(result, levels = result_levels, labels = result_labels)) |>
  count(workaround, result) |>
  group_by(workaround) |> mutate(pct = n / sum(n))
ggplot(fig5_data, aes(x = workaround, y = pct, fill = result)) +
  geom_col(width = 0.4, colour = "white", linewidth = 0.5) +
  geom_text(aes(label = if_else(pct > 0.04, percent(pct, accuracy = 1), "")),
            position = position_stack(vjust = 0.5), size = 4.5, fontface = "bold",
            colour = "white") +
  scale_fill_manual(values = result_colours) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.03))) +
  labs(title = "Per-chunk export PATH: correct diagnosis, wrong level of abstraction",
       x = NULL, y = "Proportion of chunk attempts") +
  theme_dsc()
```

## Failure Mode 3 — CRLF Line Endings

**Two line-ending conventions.** Unix/Linux marks the end of a text line
with a single newline character (`\n`, LF). Windows uses two characters:
carriage return + line feed (`\r\n`, CRLF), inherited from physical typewriter
mechanics where a carriage return moved the print head to the line start and
a line feed advanced the paper.

**How CRLF corrupts bash.** When a Windows browser downloads a text file,
Windows may save it with `\r\n` endings. Inside WSL2's Linux kernel, bash
reads the `\r` as part of the last word on each line. The chunk:

```bash
ls -d data/raw
```

becomes `ls -d data/raw\r` from bash's perspective, which it cannot find.
The characteristic symptom: `cannot: command not found` and `ls:: command
not found`, where the colon is the `\r` rendered as a visible glyph.

**Why only 6 of 18 WSL2 students were affected.** CRLF corruption requires
a Windows process to write the file. The other 12 WSL2 students obtained the
notebook entirely on the Linux side — using a Linux browser inside WSL2,
a WSL2 terminal `wget` or `cp`, or similar. When no Windows process touches
the file, no CRLF conversion occurs.

**Could an Edge user have avoided it?** Only by running `dos2unix` afterward
(unlikely without knowing the problem first), or if they happened to copy
the file from Windows to Linux using `cp` — but `cp` copies bytes as-is so
CRLF is preserved. Edge's CRLF behavior depends on Windows settings and
content-type headers and cannot be relied on in either direction.

### Detecting and Fixing CRLF

Before opening any `.qmd` file obtained through a Windows browser, verify
its line endings:

```bash
cat -A filename.qmd | head -5
```

`cat -A` makes invisible characters visible. Clean Unix line endings show
`$` at the end of each line; CRLF contamination shows `^M$`:

```
# Clean:
ls -d data/raw$

# CRLF contaminated:
ls -d data/raw^M$
```

If `^M$` appears, fix it immediately:

```bash
sudo apt install dos2unix     # WSL2/Ubuntu (one-time)
brew install dos2unix         # macOS (rarely needed)

dos2unix filename.qmd

cat -A filename.qmd | head -5  # verify: should now show $ not ^M$
```

**Prevention strategies, in order of reliability:**

1. **Use `wget` inside WSL2** — Demo 29 and all subsequent notebooks are
   delivered this way. No Windows process is involved.
2. **Use a Linux browser** — Install Firefox inside WSL2
   (`sudo apt install firefox`) and use it to access CatCourses. Downloads
   stay on the Linux filesystem.
3. **Run `dos2unix` as a precaution** — if a file arrived through any Windows
   process (Edge, Chrome on Windows, OneDrive sync), run `dos2unix` on it
   before opening in RStudio. One second of prevention prevents hours of
   debugging.
4. **Never open `.qmd` files in Windows text editors** — Notepad, Word, and
   Windows-side VS Code all write CRLF. Use editors inside WSL2 (`emacs`,
   `mg`, `nano`, or `code .` from the WSL2 terminal).

## All Failures Together — Per-Student View

```{r fig-all-students, echo=FALSE, fig.height=7}
tile_data <- students |>
  select(student, platform, all_of(chunk_levels)) |>
  pivot_longer(all_of(chunk_levels), names_to = "chunk", values_to = "result") |>
  mutate(chunk  = factor(chunk,  levels = chunk_levels, labels = chunk_labels),
         result = factor(result, levels = result_levels, labels = result_labels)) |>
  group_by(student) |> mutate(n_correct = sum(result == "CORRECT")) |> ungroup()
student_order <- tile_data |>
  distinct(student, platform, n_correct) |>
  arrange(platform, desc(n_correct), student) |> pull(student)
tile_data <- tile_data |> mutate(student = factor(student, levels = student_order))
ggplot(tile_data, aes(x = chunk, y = student, fill = result)) +
  geom_tile(colour = "white", linewidth = 0.35) +
  scale_fill_manual(values = result_colours) +
  facet_grid(rows = vars(platform), scales = "free_y", space = "free_y") +
  labs(title = "Every student's outcome on every chunk, grouped by platform",
       subtitle = "Ordered by CORRECT count within platform. All-ERR = PATH issue. All-no_output = CRLF.",
       x = NULL, y = NULL) +
  theme_dsc(base = 11) +
  theme(axis.text.y = element_text(size = 7), axis.text.x = element_text(size = 9),
        strip.text.y = element_text(size = 8, angle = 0, hjust = 0),
        panel.grid = element_blank(), legend.position = "bottom")
```

## The Two Main Failures Were Independent

```{r fig-cooccurrence, echo=FALSE, fig.height=5}
fig7_data <- students |>
  mutate(
    path_ok = if_else(!if_any(all_of(chunk_levels), \(x) x == "ERR"),
                      "check_bash\nreachable on PATH", "check_bash\nNOT reachable"),
    cwd_ok  = if_else(correct_cwd, "Correct CWD", "Wrong CWD"),
    path_ok = factor(path_ok, levels = c("check_bash\nreachable on PATH",
                                         "check_bash\nNOT reachable")),
    cwd_ok  = factor(cwd_ok,  levels = c("Correct CWD","Wrong CWD"))
  ) |>
  count(cwd_ok, path_ok) |>
  mutate(pct = round(n/n_total*100),
         fill_grp = paste(as.character(cwd_ok), as.character(path_ok)))
tile_fills <- c(
  "Correct CWD check_bash\nreachable on PATH" = clr_correct,
  "Correct CWD check_bash\nNOT reachable"     = clr_incorrect,
  "Wrong CWD check_bash\nreachable on PATH"   = clr_noout,
  "Wrong CWD check_bash\nNOT reachable"       = clr_err
)
ggplot(fig7_data, aes(x = path_ok, y = fct_rev(cwd_ok), fill = fill_grp)) +
  geom_tile(colour = "white", linewidth = 2) +
  geom_text(aes(label = paste0(n, " students\n(", pct, "%)")),
            size = 6, fontface = "bold", colour = "white", lineheight = 1.4) +
  scale_fill_manual(values = tile_fills, guide = "none") +
  labs(title = "Fixing one problem does not fix the other — both had to be addressed",
       subtitle = "14 had only PATH issue · 10 had only CWD issue · 19 had both · 9 had neither",
       x = NULL, y = NULL,
       caption = paste0(
         "PATH fix: check_bash 2.2 to /usr/local/bin/\n",
         "CWD fix:  error=FALSE guard chunk halts rendering if path is wrong")) +
  theme_dsc() + theme(panel.grid = element_blank(), axis.text = element_text(size = 12))
```

## What Changed for Demo 29 and Lab 10

**Problem 1 — `check_bash` not on PATH (63% of students)**

`check_bash` 2.2 is installed to `/usr/local/bin/` via `sudo cp` from a
cloned repository. `/usr/local/bin/` is on the default PATH of every bash
subprocess on both platforms, regardless of how RStudio was launched. No
per-chunk export, no setup-chunk PATH manipulation, and no `engine.path`
setting is needed for `check_bash` to be found.

*General principle:* Tools needed inside knitr chunks must be installed
system-wide (`sudo` to `/usr/local/bin/`) or via an environment manager
like `pixi` that places binaries on the activated PATH. The `~/bin/` pattern
works only for interactive terminal sessions.

**Problem 2 — Wrong working directory (56% of students)**

Demo 29 opens with a CWD guard using `error=FALSE`:

```bash
expected="$HOME/DSC-011-S26/DSC-011-S26-Baby-Names"
actual=$(pwd)
if [[ "$actual" != "$expected" ]]; then
    echo "ERROR: Working directory is $actual"
    echo "       Expected: $expected"
    exit 1
fi
```

`error=FALSE` makes knitr treat a non-zero bash exit as a fatal rendering
error — the notebook stops immediately with a clear diagnostic message.

**Problem 3 — macOS using `/bin/bash` 3.2 (12% of students)**

The bash version guard now uses `error=FALSE`. The R setup chunk sets
`engine.path = list(bash = "/opt/homebrew/bin/bash")` when that path exists.
macOS students are instructed to launch RStudio from the terminal:
`open -a RStudio`

**Problem 4 — CRLF line endings (12% of students)**

All notebooks from Demo 29 onward are delivered via `wget` from a course URL,
keeping the file entirely on the Linux filesystem. Emergency fix: `dos2unix`.

**Problems 5 and 6 — `ls` under `set -euo pipefail`; stale `VIRTUAL_ENV`**

Both were notebook design bugs. `ls` was replaced with `find`. 
`Sys.unsetenv("VIRTUAL_ENV")` was added to the R setup chunk.

## Lessons Learned: Habits and Practices

The infrastructure fixes above address the technical causes. The following
require sustained attention across the remaining course sessions.

### Never type directory or file names by hand

`r n_other_wrong` students had paths with typos or near-misses in the project
or course name: `DSC-011` vs `DSC011`, `DCS_011` vs `DSC_011`, nested in a
`Lecture28/` subfolder, or at the parent directory rather than the project
root. A path differing by one character is a different path.

**Do not type long paths. Instead:**

- **Copy-paste** from course materials.
- **Tab-complete** in the terminal: type the first few characters and press
  <kbd>Tab</kbd>. If nothing completes, the directory does not exist at that
  location — this is a diagnostic, not a typing problem.

### Fix the clipboard before every session

Copy-paste is the recommended alternative to typing, but it requires setup
on some WSL2 configurations.

**macOS:** <kbd>Cmd</kbd>+<kbd>C</kbd> / <kbd>Cmd</kbd>+<kbd>V</kbd>. Works
natively.

**WSL2 — Windows Terminal (recommended):** The
<kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>C</kbd> /
<kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>V</kbd> shortcuts are **not enabled by
default**. A student in this course identified and documented this issue. To
enable them:

1. Right-click the Windows Terminal title bar.
2. Select **Properties**.
3. Check **"Use Ctrl+Shift+C/V as Copy/Paste"**.
4. Click OK.

Right-click also pastes in most Windows Terminal configurations. **Verify
copy-paste works before every session** — paste a long path from course
materials into the terminal and confirm it arrives correctly.

**WSL2 — standalone Linux terminal (WSLg, xterm, GNOME Terminal):** The
clipboard bridge may not be active by default. Install clipboard tools:

```bash
sudo apt install xclip        # X11-based terminals
sudo apt install wl-clipboard # Wayland-based WSLg
```

If pasting still fails, switch to Windows Terminal, which has reliable
clipboard integration.

### Do not work from Windows filesystem paths in WSL2

`r n_windows_path` students (15%) rendered from `/mnt/c/` or `/c/` paths.
If the terminal prompt shows a path beginning with `/mnt/c/`, the working
location is wrong. All course work lives in `~/DSC-011-S26/` on the Linux
filesystem.

### Use a Linux browser for CatCourses in WSL2

The CRLF problem was caused by downloading through a Windows browser.
Install Firefox inside WSL2 and use it for CatCourses:

```bash
sudo apt install firefox
firefox &   # launch from inside WSL2
```

Run `dos2unix` on any file obtained through a Windows process as a
precaution, and verify with `cat -A`.

### Keep the project in the course-standard location

`r n_downloads + n_desktop` students were in Downloads or Desktop.
`r n_other_wrong` used non-standard names or locations. The course standard:

```
~/DSC-011-S26/
└── DSC-011-S26-Baby-Names/
    ├── data/raw/
    ├── data/processed/
    ├── data/clean/
    └── scripts/
```

Create this structure once with `mkdir -p` at the start of Demo 29. The CWD
guard in every subsequent notebook will halt rendering if the project is
anywhere else.

## Summary

This document demonstrates standard reproducible data science practice: a
self-contained analysis starting from raw data, with every step documented,
anonymization machine-verifiable, and conclusions traceable to the data.

The Demo 28 failures were caused by **infrastructure problems, not student
capability problems.** `check_bash` worked correctly when found. Students who
had the tool on PATH and the correct working directory passed validation chunks.
The three infrastructure fixes — system-level `check_bash` install,
`error=FALSE` guard chunks, and `wget` notebook delivery — remove these
problems permanently. The five habit-level lessons require sustained attention.

```{r session-info, echo=FALSE}
sessionInfo()
```
