---
title: "Demo 28: What the Data Showed"
subtitle: "DSC 011 Sp26 · Submission Analysis (anonymized) · n = 52 (Linux/UNIX Project Building and Infrastructure Check)"
author: "David H. Ardell with Claude Sonnet 4.6"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    fig-align: center
    fig-width: 10
    fig-height: 5.8
    code-fold: true
    highlight-style: github
    footer: "DSC 011 Sp26 — Demo 28 Submission Analysis (anonymized)"
    css: |
      .reveal .slides section {
        font-size: 0.82em;
        line-height: 1.5;
      }
      .reveal h2 { font-size: 1.2em; margin-bottom: 0.4em; }
      .reveal pre { font-size: 0.72em; }
      .caption-slide p { font-size: 0.78em; color: #444; }
execute:
  echo: true
  warning: false
  message: false
engine: knitr
---

```{r setup, include=FALSE}
options(digits = 5)
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, warning = FALSE,
                      message = FALSE, comment = "")
```

```{r packages, include=FALSE}
library(ggplot2); library(dplyr); library(tidyr)
library(forcats);  library(scales); library(stringr)
```

```{r palette, include=FALSE}
OI <- c(orange="#E69F00", sky_blue="#56B4E9", green="#009E73",
        yellow="#F0E442", blue="#0072B2", vermillion="#D55E00",
        pink="#CC79A7",   black="#000000")
clr_correct <- OI[["green"]];  clr_incorrect <- OI[["orange"]]
clr_err     <- OI[["vermillion"]]; clr_noout  <- OI[["sky_blue"]]
result_colours <- c("CORRECT"=clr_correct,"INCORRECT"=clr_incorrect,
                    "Error running bash"=clr_err,"No output / not run"=clr_noout)
plat_colours <- c("macOS (Homebrew bash 5)"=OI[["blue"]],
                  "WSL2/Linux (bash 5.2)"=OI[["green"]],
                  "macOS (/bin/bash 3.2, unfixed)"=OI[["pink"]],
                  "WSL2 (CRLF line endings)"=OI[["orange"]])
cwd_colours <- c("Correct"=OI[["green"]],
                 "Windows path (/mnt/c/ or /c/)"=OI[["vermillion"]],
                 "Downloads/"=OI[["orange"]],"Desktop/"=OI[["pink"]],
                 "Other wrong location"=OI[["sky_blue"]])
theme_dsc <- function(base=13) {
  theme_minimal(base_size=base) +
    theme(plot.title=element_text(face="bold",size=base+1,margin=margin(b=4)),
          plot.subtitle=element_text(size=base-1,color="grey30",lineheight=1.25,
                                     margin=margin(b=6)),
          plot.caption=element_text(size=base-3,color="grey50",hjust=0,
                                    lineheight=1.3,margin=margin(t=6)),
          legend.position="bottom", legend.title=element_blank(),
          legend.text=element_text(size=base-2),
          panel.grid.minor=element_blank(), plot.margin=margin(4,4,4,4))
}
```

```{r data, include=FALSE}
raw <- tibble(
  student = paste0("student_", 1:52),
  late = c(TRUE,FALSE,FALSE,TRUE,FALSE,TRUE,TRUE,FALSE,TRUE,FALSE,TRUE,FALSE,
            TRUE,FALSE,TRUE,FALSE,FALSE,TRUE,FALSE,TRUE,FALSE,TRUE,TRUE,FALSE,
            TRUE,FALSE,FALSE,TRUE,FALSE,FALSE,TRUE,TRUE,FALSE,FALSE,FALSE,TRUE,
            TRUE,TRUE,FALSE,TRUE,FALSE,TRUE,FALSE,FALSE,FALSE,TRUE,FALSE,TRUE,
            FALSE,TRUE,FALSE,TRUE),
  platform = c(
    "macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)",
    "WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)",
    "macOS (Homebrew bash 5)","macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)",
    "macOS (Homebrew bash 5)","WSL2 (CRLF line endings)","macOS (Homebrew bash 5)",
    "macOS (/bin/bash 3.2, unfixed)","WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)",
    "WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)",
    "macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)","macOS (/bin/bash 3.2, unfixed)",
    "WSL2 (CRLF line endings)","macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)",
    "macOS (/bin/bash 3.2, unfixed)","WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)",
    "WSL2/Linux (bash 5.2)","WSL2 (CRLF line endings)","macOS (Homebrew bash 5)",
    "macOS (Homebrew bash 5)","macOS (/bin/bash 3.2, unfixed)","WSL2/Linux (bash 5.2)",
    "macOS (Homebrew bash 5)","macOS (Homebrew bash 5)","WSL2 (CRLF line endings)",
    "WSL2/Linux (bash 5.2)","macOS (/bin/bash 3.2, unfixed)","macOS (Homebrew bash 5)",
    "WSL2/Linux (bash 5.2)","WSL2 (CRLF line endings)","macOS (Homebrew bash 5)",
    "WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)",
    "macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)","macOS (Homebrew bash 5)",
    "WSL2/Linux (bash 5.2)","macOS (/bin/bash 3.2, unfixed)","WSL2/Linux (bash 5.2)",
    "WSL2/Linux (bash 5.2)"),
  correct_cwd = c(TRUE,FALSE,TRUE,TRUE,TRUE,FALSE,TRUE,FALSE,TRUE,FALSE,FALSE,
                  FALSE,FALSE,TRUE,TRUE,FALSE,FALSE,TRUE,FALSE,TRUE,FALSE,FALSE,
                  FALSE,TRUE,FALSE,FALSE,TRUE,TRUE,FALSE,FALSE,TRUE,FALSE,TRUE,
                  FALSE,FALSE,TRUE,TRUE,FALSE,TRUE,FALSE,FALSE,TRUE,TRUE,TRUE,
                  FALSE,TRUE,FALSE,FALSE,TRUE,FALSE,TRUE,FALSE),
  cwd_category = c(
    "Correct","Desktop/","Correct","Correct","Correct","Other wrong location",
    "Correct","Windows path (/mnt/c/ or /c/)","Correct","Downloads/",
    "Windows path (/mnt/c/ or /c/)","Desktop/","Downloads/","Correct","Desktop/",
    "Correct","Downloads/","Correct","Downloads/","Correct","Other wrong location",
    "Windows path (/mnt/c/ or /c/)","Other wrong location","Correct","Downloads/",
    "Correct","Correct","Correct","Windows path (/mnt/c/ or /c/)","Downloads/",
    "Other wrong location","Downloads/","Correct","Downloads/","Desktop/",
    "Windows path (/mnt/c/ or /c/)","Correct","Other wrong location","Correct",
    "Other wrong location","Windows path (/mnt/c/ or /c/)","Correct",
    "Other wrong location","Desktop/","Other wrong location","Correct",
    "Other wrong location","Downloads/","Correct","Windows path (/mnt/c/ or /c/)",
    "Other wrong location","Windows path (/mnt/c/ or /c/)"),
  export_fix = c(FALSE,FALSE,FALSE,TRUE,FALSE,FALSE,FALSE,FALSE,TRUE,FALSE,FALSE,
                 FALSE,FALSE,FALSE,FALSE,TRUE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,
                 FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,TRUE,FALSE,FALSE,
                 FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,TRUE,FALSE,FALSE,FALSE,FALSE,
                 FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,TRUE,FALSE),
  dir_scaffold = c("ERR","ERR","CORRECT","ERR","CORRECT","ERR","no_output","no_output",
    "no_output","no_output","no_output","no_output","ERR","no_output","INCORRECT",
    "ERR","ERR","no_output","ERR","no_output","ERR","no_output","INCORRECT","ERR",
    "ERR","ERR","no_output","no_output","no_output","no_output","no_output","ERR",
    "ERR","ERR","ERR","no_output","ERR","ERR","no_output","ERR","no_output",
    "CORRECT","no_output","ERR","ERR","no_output","ERR","ERR","ERR","no_output",
    "no_output","no_output"),
  python_env = c("ERR","ERR","CORRECT","ERR","ERR","ERR","INCORRECT","no_output",
    "ERR","no_output","no_output","no_output","ERR","ERR","INCORRECT","ERR","ERR",
    "ERR","ERR","no_output","ERR","no_output","INCORRECT","ERR","ERR","ERR",
    "CORRECT","ERR","no_output","no_output","ERR","ERR","ERR","ERR","ERR",
    "no_output","ERR","ERR","CORRECT","ERR","no_output","CORRECT","ERR","ERR",
    "ERR","no_output","ERR","ERR","ERR","no_output","ERR","no_output"),
  setup_sh = c("ERR","ERR","CORRECT","ERR","ERR","ERR","INCORRECT","no_output",
    "ERR","no_output","no_output","INCORRECT","ERR","ERR","INCORRECT","ERR","ERR",
    "ERR","ERR","INCORRECT","ERR","no_output","INCORRECT","ERR","ERR","ERR","ERR",
    "ERR","no_output","no_output","ERR","ERR","ERR","ERR","ERR","no_output","ERR",
    "ERR","CORRECT","ERR","no_output","INCORRECT","ERR","ERR","ERR","INCORRECT",
    "ERR","ERR","ERR","no_output","ERR","no_output"),
  git_commit_1 = c("ERR","ERR","INCORRECT","ERR","ERR","ERR","INCORRECT","no_output",
    "ERR","no_output","no_output","INCORRECT","ERR","ERR","INCORRECT","ERR","ERR",
    "ERR","ERR","INCORRECT","ERR","no_output","INCORRECT","ERR","ERR","ERR","ERR",
    "ERR","no_output","no_output","ERR","ERR","ERR","ERR","ERR","no_output","ERR",
    "ERR","INCORRECT","ERR","no_output","INCORRECT","ERR","ERR","ERR","INCORRECT",
    "ERR","ERR","ERR","no_output","ERR","no_output"),
  html_14_files = c("ERR","ERR","INCORRECT","ERR","ERR","ERR","INCORRECT","no_output",
    "ERR","no_output","no_output","INCORRECT","ERR","ERR","INCORRECT","ERR","ERR",
    "ERR","ERR","INCORRECT","ERR","no_output","INCORRECT","ERR","ERR","ERR","ERR",
    "ERR","no_output","no_output","ERR","ERR","ERR","ERR","ERR","no_output","ERR",
    "ERR","CORRECT","ERR","no_output","INCORRECT","ERR","ERR","ERR","INCORRECT",
    "ERR","ERR","ERR","no_output","ERR","no_output"),
  git_commit_2 = c("ERR","ERR","INCORRECT","ERR","ERR","ERR","INCORRECT","no_output",
    "ERR","no_output","no_output","INCORRECT","ERR","ERR","INCORRECT","ERR","ERR",
    "ERR","ERR","INCORRECT","ERR","no_output","INCORRECT","ERR","ERR","ERR","ERR",
    "ERR","no_output","no_output","ERR","ERR","ERR","ERR","ERR","no_output","ERR",
    "ERR","CORRECT","ERR","no_output","CORRECT","ERR","ERR","ERR","INCORRECT",
    "ERR","ERR","ERR","no_output","ERR","no_output")
)
n_total <- nrow(raw)
chunk_levels <- c("dir_scaffold","python_env","setup_sh",
                  "git_commit_1","html_14_files","git_commit_2")
chunk_labels <- c(dir_scaffold="Dir scaffold", python_env="Python env",
                  setup_sh="00-setup.sh", git_commit_1="Git commits = 1",
                  html_14_files="14 HTML files", git_commit_2="Git commits = 2")
result_levels <- c("CORRECT","INCORRECT","ERR","no_output")
result_labels  <- c("CORRECT","INCORRECT","Error running bash","No output / not run")
plat_levels <- c("macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)",
                 "macOS (/bin/bash 3.2, unfixed)","WSL2 (CRLF line endings)")
cwd_cat_levels <- c("Correct","Windows path (/mnt/c/ or /c/)",
                    "Downloads/","Desktop/","Other wrong location")
students <- raw |>
  mutate(platform     = factor(platform,     levels = plat_levels),
         cwd_category = factor(cwd_category, levels = cwd_cat_levels),
         submission   = if_else(late, "Late","On time"))
n_windows_path <- sum(students$cwd_category == "Windows path (/mnt/c/ or /c/)")
n_downloads    <- sum(students$cwd_category == "Downloads/")
n_desktop      <- sum(students$cwd_category == "Desktop/")
n_other_wrong  <- sum(students$cwd_category == "Other wrong location")
n_correct_cwd  <- sum(students$correct_cwd)
n_wrong_cwd    <- n_total - n_correct_cwd
```

---

## Three failure modes, independently caused

Every Demo 28 error traces back to one or more of three root causes:

1. **`check_bash` not on PATH** in knitr bash subprocesses → 33 students (63%)
2. **Wrong working directory** — notebook not in the project root → 29 students (56%)
3. **Windows CRLF line endings** in the `.qmd` file → 6 students (12%)

**Avoid Causes 2 and 3: follow instructions!**

---

## The class split into four platform groups

Two OS × two configuration states. The configuration problems are the story.

```{r fig-platform, echo=FALSE}
fig1_data <- students |> count(platform, submission)
ggplot(fig1_data, aes(x=submission, y=n, fill=platform)) +
  geom_col(width=0.45, colour="white", linewidth=0.6) +
  geom_text(aes(label=n), position=position_stack(vjust=0.5),
            size=5, fontface="bold", colour="white") +
  scale_fill_manual(values=plat_colours) +
  scale_y_continuous(expand=expansion(mult=c(0,0.06))) +
  labs(title="Two groups worked; two reflect misconfiguration",
       subtitle=paste0(
         "macOS /bin/bash 3.2: RStudio launched from Dock, not terminal  ·  ",
         "WSL2 CRLF: .qmd downloaded through Windows browser"),
       x=NULL, y="Students",
       caption="n = 52  ·  22 macOS Homebrew, 18 WSL2, 6 macOS unfixed, 6 WSL2 CRLF") +
  theme_dsc()
```

---

## macOS (/bin/bash 3.2): silent failure

When RStudio is launched from the Dock, macOS's process manager (`launchd`)
gives it a minimal PATH that excludes `/opt/homebrew/bin/`. knitr falls back
to `/bin/bash` — Apple's GPL-frozen bash 3.2 from 2007.

The bash version guard was designed to catch this. But Demo 28 used
`error=TRUE`, so knitr **swallowed the non-zero exit silently** and continued
into all subsequent chunks with a broken environment.

**Demo 29 fix:** bash version guard now uses `error=FALSE` — rendering halts
immediately. macOS students instructed to launch RStudio with `open -a RStudio`.

---

## Failure Mode 1 — Wrong working directory

When Quarto renders, knitr sets the working directory to the folder containing
the `.qmd` file. Bash chunks use that as their current directory. Relative
paths like `data/raw/` and `scripts/` only resolve from the project root.

`r n_wrong_cwd` of `r n_total` students (**`r round(n_wrong_cwd/n_total*100)`%**)
rendered from the wrong location.

**Demo 29 fix:** `error=FALSE` CWD guard halts rendering if the path ≠
`~/DSC-011-S26/DSC-011-S26-Baby-Names`.

---

## Where students actually were

```{r fig-cwd, echo=FALSE}
cwd_data <- students |> count(cwd_category) |> mutate(cwd_category=fct_rev(cwd_category))
ggplot(cwd_data, aes(x=cwd_category, y=n, fill=cwd_category)) +
  geom_col(width=0.55, colour="white", linewidth=0.5) +
  geom_text(aes(label=paste0(n," (",round(n/n_total*100),"%)")),
            hjust=-0.1, size=4.5, fontface="bold", colour="grey20") +
  scale_fill_manual(values=cwd_colours, guide="none") +
  scale_y_continuous(expand=expansion(mult=c(0,0.35))) +
  coord_flip() +
  labs(title=paste0(n_wrong_cwd," of ",n_total," students (",
                    round(n_wrong_cwd/n_total*100),"%) in the wrong directory"),
       subtitle=paste0(
         n_windows_path," in Windows paths  ·  ",
         n_downloads," in Downloads/  ·  ",
         n_desktop," on Desktop/  ·  ",
         n_other_wrong," in other wrong locations"),
       x=NULL, y="Students",
       caption="The variety of wrong paths shows this was not one mistake — many different ones.") +
  theme_dsc()
```

---

## Windows paths: a distinct sub-problem

**`r n_windows_path` students (15%) worked from `/mnt/c/` or `/c/` paths**
— the Windows NTFS filesystem mounted inside WSL2.

Working from `/mnt/c/` causes:

- **CRLF contamination** — Windows tools write `\r\n` line endings that break bash
- **Slow filesystem I/O** — cross-boundary overhead on every read/write
- **Unreliable `chmod`** — Linux executable bits don't map to NTFS permissions
- **Non-portability** — paths are specific to one machine's Windows layout

**All course work lives in `~/DSC-011-S26/` on the Linux filesystem.**
If the terminal shows a `/mnt/c/` path: wrong location.

---

## Directory and file names must be exact

`r n_other_wrong` students had near-miss paths: `DSC-011` vs `DSC011`,
`DCS_011` vs `DSC_011`, nested inside `Lecture28/`, or at the parent
directory instead of the project root.

A path that differs by one character is a different path.

**Never type long directory or file names by hand.**

- **Copy-paste** from course materials
- **Tab-complete** in the terminal — if nothing completes, the directory
  does not exist at that location (diagnostic, not a typing problem)

---

## Fix copy-paste before every session (WSL2)

Clipboard shortcuts are **not enabled by default** in Windows Terminal.
A student in this course identified and documented this fix:

**Windows Terminal (recommended):**

1. Right-click the title bar → **Properties**
2. Check **"Use Ctrl+Shift+C/V as Copy/Paste"**
3. Click OK

Then use <kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>C</kbd> /
<kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>V</kbd>. Right-click also pastes.

**Standalone Linux terminal (WSLg/xterm):**

```bash
sudo apt install xclip          # X11
sudo apt install wl-clipboard   # Wayland/WSLg
```

If pasting still fails: switch to Windows Terminal.

---

## Failure Mode 2 — `check_bash` not on PATH

`check_bash` was installed to `~/bin/` via curl. The installer added
`export PATH="$HOME/bin:$PATH"` to `.bashrc`. This works in an
**interactive** terminal session.

knitr bash chunks are **non-interactive** bash subprocesses. They do not
source `.bashrc` or `.zshrc`. They inherit only R's environment at render
time — which, on both macOS and WSL2, does not include `~/bin/`.

Result: `check_bash` worked in the terminal. Every knitr chunk calling it
got `bash: check_bash: command not found` (exit 127), rendered by knitr
as `Error in running command bash`.

---

## 0 students passed the first validation chunk

```{r fig-chunks, echo=FALSE}
fig3_data <- students |>
  select(student, all_of(chunk_levels)) |>
  pivot_longer(-student, names_to="chunk", values_to="result") |>
  mutate(chunk  = factor(chunk,  levels=chunk_levels, labels=chunk_labels),
         result = factor(result, levels=result_levels, labels=result_labels)) |>
  count(chunk, result)
ggplot(fig3_data, aes(x=chunk, y=n, fill=result)) +
  geom_col(width=0.55, colour="white", linewidth=0.5) +
  geom_text(aes(label=n), position=position_stack(vjust=0.5),
            size=4.8, fontface="bold", colour="white") +
  scale_fill_manual(values=result_colours) +
  scale_y_continuous(expand=expansion(mult=c(0,0.04)), breaks=seq(0,52,10)) +
  labs(title="0 CORRECT on dir scaffold · 33 students had bash errors on every chunk",
       subtitle=paste0(
         "'Error running bash' = check_bash not found (exit 127)  ·  ",
         "'No output' = chunk skipped or CRLF  ·  ",
         "'INCORRECT' = check_bash ran, answer wrong"),
       x=NULL, y="Students",
       caption="Demo 29 fix: check_bash 2.2 installed to /usr/local/bin/ — on PATH everywhere by default.") +
  theme_dsc() + theme(axis.text.x=element_text(size=10))
```

---

## The PATH problem hit macOS and WSL2 equally

```{r fig-heatmap, echo=FALSE, fig.height=4.5}
fig4_data <- students |>
  filter(platform %in% c("macOS (Homebrew bash 5)","WSL2/Linux (bash 5.2)")) |>
  select(student, platform, all_of(chunk_levels)) |>
  pivot_longer(all_of(chunk_levels), names_to="chunk", values_to="result") |>
  mutate(chunk=factor(chunk, levels=chunk_levels, labels=chunk_labels),
         is_err=result=="ERR") |>
  group_by(platform, chunk) |>
  summarise(n_err=sum(is_err), n=n(), pct=n_err/n, .groups="drop")
ggplot(fig4_data, aes(x=chunk, y=fct_rev(platform), fill=pct)) +
  geom_tile(colour="white", linewidth=1.2) +
  geom_text(aes(label=paste0(n_err,"/",n,"\n(",round(pct*100),"%)")),
            size=3.8, lineheight=1.2) +
  scale_fill_gradient(low="#fffde7", high=clr_err,
                      labels=percent_format(accuracy=1),
                      name="% with error", limits=c(0,1)) +
  labs(title="macOS and WSL2 failed at nearly identical rates",
       subtitle=paste0(
         "WSL2 RStudio Server also does not source .bashrc when spawning knitr subprocesses.\n",
         "This was an architectural problem, not a macOS GUI PATH problem."),
       x=NULL, y=NULL,
       caption="/usr/local/bin/ is on the default PATH of every bash subprocess on both platforms.") +
  theme_dsc() + theme(panel.grid=element_blank(), legend.position="right",
                      axis.text.x=element_text(size=10))
```

---

## The fix: install to a system-level directory

The fundamental principle: tools that must be available to **non-interactive**
subprocesses must be in a directory on the **default PATH** of every process.

| Location | Available in interactive terminal? | Available in knitr bash chunks? |
|:---------|:-:|:-:|
| `~/bin/` | ✓ (if `.bashrc` sourced) | ✗ |
| `/usr/local/bin/` | ✓ | ✓ |

This principle applies to any tool needed inside knitr chunks. Install
system-wide with `sudo`, or use an environment manager like `pixi` that
places binaries on the activated PATH.

**Demo 29 fix:** `check_bash` 2.2 installed to `/usr/local/bin/` via
`sudo cp`. No per-chunk export, no setup-chunk PATH patch needed.

---

## Six students found the per-chunk workaround

```{r fig-workaround, echo=FALSE, fig.height=4.8}
fig5_data <- students |>
  mutate(workaround=if_else(export_fix,
    paste0("Used export PATH workaround (n=",sum(export_fix),")"),
    paste0("Did not (n=",sum(!export_fix),")"))) |>
  select(student, workaround, all_of(chunk_levels)) |>
  pivot_longer(all_of(chunk_levels), names_to="chunk", values_to="result") |>
  mutate(result=factor(result, levels=result_levels, labels=result_labels)) |>
  count(workaround, result) |>
  group_by(workaround) |> mutate(pct=n/sum(n))
ggplot(fig5_data, aes(x=workaround, y=pct, fill=result)) +
  geom_col(width=0.4, colour="white", linewidth=0.5) +
  geom_text(aes(label=if_else(pct>0.04, percent(pct,accuracy=1),"")),
            position=position_stack(vjust=0.5),
            size=4.5, fontface="bold", colour="white") +
  scale_fill_manual(values=result_colours) +
  scale_y_continuous(labels=percent_format(accuracy=1),
                     expand=expansion(mult=c(0,0.03))) +
  labs(title="Correct diagnosis — wrong level of abstraction",
       subtitle=paste0(
         "Adding export PATH to some chunks improved the rate but not enough.\n",
         "Applied inconsistently: some chunks fixed, others still errored."),
       x=NULL, y="Proportion of chunk attempts",
       caption="The fix belongs at installation time, not inside each chunk.") +
  theme_dsc()
```

---

## Failure Mode 3 — CRLF line endings (WSL2 only)

Unix/Linux uses `\n` (LF) to end lines. Windows uses `\r\n` (CRLF).
When a Windows browser downloads a text file, Windows may save it with
`\r\n` endings.

Inside WSL2's Linux kernel, bash reads the `\r` as part of the last word
on each line. The chunk `ls -d data/raw` becomes `ls -d data/raw\r` —
which bash cannot find. Symptom: `cannot: command not found` and
`ls:: command not found` (the colon is `\r` rendered as a glyph).

**Why only 6 of 18 WSL2 students?** The other 12 obtained the notebook
through a Linux process — a Linux browser inside WSL2, or `wget` from the
terminal. When no Windows process touches the file, no CRLF occurs.

---

## Could an Edge user have avoided CRLF?

Only in two ways:

1. **Run `dos2unix` explicitly** before opening the file — but this requires
   knowing the problem exists before seeing the error. Unlikely.

2. **`cp` from `/mnt/c/` into `~/`** — does **not** fix CRLF. `cp` copies
   bytes as-is; the `\r\n` endings travel with the file.

Edge's CRLF behavior depends on Windows settings and content-type headers —
**it cannot be relied upon in either direction.**

The only reliable approaches: (1) never let a Windows process write the file,
or (2) always run `dos2unix` on any file from a Windows source.

---

## Detecting and fixing CRLF

**Detect** before opening in RStudio:

```bash
cat -A filename.qmd | head -5
```

Clean file shows `$` at line ends. CRLF file shows `^M$`:

```
ls -d data/raw$      ← clean
ls -d data/raw^M$    ← CRLF contaminated
```

**Fix immediately:**

```bash
sudo apt install dos2unix    # WSL2 (one-time)
dos2unix filename.qmd
cat -A filename.qmd | head -5   # verify: should show $, not ^M$
```

---

## Prevention: keep files on the Linux side

**In order of reliability:**

1. **`wget` inside WSL2** — Demo 29+ all delivered this way. A Linux process
   downloads directly; no Windows involvement.
2. **Linux browser for CatCourses** — `sudo apt install firefox`, launch with
   `firefox &`. Downloads stay on the Linux filesystem.
3. **`dos2unix` as precaution** — run on any file from Edge, Chrome on
   Windows, or OneDrive sync before opening in RStudio.
4. **Never edit `.qmd` files in Windows-side editors** — Notepad, Word, and
   Windows-side VS Code all write CRLF. Use `emacs`, `mg`, `nano`, or
   `code .` launched from *inside* WSL2.

---

## Every student's outcome, all at once

```{r fig-all, echo=FALSE, fig.height=6.2}
tile_data <- students |>
  select(student, platform, all_of(chunk_levels)) |>
  pivot_longer(all_of(chunk_levels), names_to="chunk", values_to="result") |>
  mutate(chunk  = factor(chunk,  levels=chunk_levels, labels=chunk_labels),
         result = factor(result, levels=result_levels, labels=result_labels)) |>
  group_by(student) |> mutate(n_correct=sum(result=="CORRECT")) |> ungroup()
student_order <- tile_data |>
  distinct(student, platform, n_correct) |>
  arrange(platform, desc(n_correct), student) |> pull(student)
tile_data <- tile_data |> mutate(student=factor(student, levels=student_order))
ggplot(tile_data, aes(x=chunk, y=student, fill=result)) +
  geom_tile(colour="white", linewidth=0.35) +
  scale_fill_manual(values=result_colours) +
  facet_grid(rows=vars(platform), scales="free_y", space="free_y") +
  labs(title="Per-student, per-chunk outcomes — grouped by platform",
       subtitle="Ordered by CORRECT count within platform. All-ERR = PATH. All-no_output = CRLF.",
       x=NULL, y=NULL) +
  theme_dsc(base=11) +
  theme(axis.text.y=element_text(size=6.5), axis.text.x=element_text(size=9),
        strip.text.y=element_text(size=8, angle=0, hjust=0),
        panel.grid=element_blank(), legend.position="bottom")
```

---

## The two main failures were independent

```{r fig-cooccurrence, echo=FALSE, fig.height=5.2}
fig7_data <- students |>
  mutate(
    path_ok=if_else(!if_any(all_of(chunk_levels),\(x)x=="ERR"),
                    "check_bash\nreachable","check_bash\nNOT reachable"),
    cwd_ok=if_else(correct_cwd,"Correct CWD","Wrong CWD"),
    path_ok=factor(path_ok, levels=c("check_bash\nreachable","check_bash\nNOT reachable")),
    cwd_ok =factor(cwd_ok,  levels=c("Correct CWD","Wrong CWD"))
  ) |>
  count(cwd_ok, path_ok) |>
  mutate(pct=round(n/n_total*100),
         fill_grp=paste(as.character(cwd_ok), as.character(path_ok)))
tile_fills <- c(
  "Correct CWD check_bash\nreachable"     = clr_correct,
  "Correct CWD check_bash\nNOT reachable" = OI[["orange"]],
  "Wrong CWD check_bash\nreachable"       = OI[["sky_blue"]],
  "Wrong CWD check_bash\nNOT reachable"   = clr_err
)
ggplot(fig7_data, aes(x=path_ok, y=fct_rev(cwd_ok), fill=fill_grp)) +
  geom_tile(colour="white", linewidth=2) +
  geom_text(aes(label=paste0(n," students\n(",pct,"%)")),
            size=6, fontface="bold", colour="white", lineheight=1.4) +
  scale_fill_manual(values=tile_fills, guide="none") +
  labs(title="Fixing one did not fix the other — both needed independent fixes",
       subtitle="14 had only PATH · 10 had only CWD · 19 had both · 9 had neither",
       x=NULL, y=NULL,
       caption=paste0(
         "PATH fix: check_bash 2.2 → /usr/local/bin/\n",
         "CWD fix:  error=FALSE guard chunk")) +
  theme_dsc() + theme(panel.grid=element_blank(), axis.text=element_text(size=12))
```

---

## What changed for Demo 29 and Lab 10 {.smaller}

| Problem | Students | Fix |
|:--------|:--------:|:----|
| `check_bash` not on knitr PATH | 33 (63%) | Install to `/usr/local/bin/` via `sudo` |
| Wrong working directory | 29 (56%) | `error=FALSE` CWD guard chunk |
| macOS using `/bin/bash` 3.2 | 6 (12%) | `error=FALSE` bash version guard + `engine.path` + `open -a RStudio` |
| CRLF line endings | 6 (12%) | `wget` delivery; `dos2unix` as rescue |
| `ls` under `set -euo pipefail` | all | Replace `ls` with `find` |
| Stale `VIRTUAL_ENV` | some | `Sys.unsetenv("VIRTUAL_ENV")` in setup chunk |

General principle: **tools needed in non-interactive subprocesses must be in
a directory on the default PATH** — `/usr/local/bin/` with `sudo`, or an
environment manager like `pixi`.

---

## Five habits for the rest of the course {.smaller}

1. **Never type long directory or file names** — copy-paste or tab-complete.
   A one-character typo is a different path.

2. **Fix copy-paste first** — Windows Terminal: right-click title bar →
   Properties → check "Use Ctrl+Shift+C/V". Test before every session.

3. **Work in `~/DSC-011-S26/` on the Linux filesystem** — never `/mnt/c/`.
   If the prompt shows `/mnt/c/`, stop and `cd ~` first.

4. **Use a Linux browser for CatCourses** (`sudo apt install firefox`) or
   download via `wget`. Run `dos2unix` on any file from a Windows source.
   Verify with `cat -A file.qmd | head -5`.

5. **Keep the project at the course-standard path** —
   `~/DSC-011-S26/DSC-011-S26-Baby-Names/`. Create once with `mkdir -p`.
   Never move it. Every notebook's CWD guard enforces this.

---

## The bottom line {.center}

These were **infrastructure failures, not capability failures.**

`check_bash` worked correctly when found.
Students who had the correct environment passed validation chunks.

The three infrastructure fixes make these errors permanently impossible.
The five habits make the infrastructure reliable.

```{r session-info, echo=FALSE, include=FALSE}
sessionInfo()
```
